Tree size, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest

### 2019-09-19 manuscript/climate_sensitivity_draft.Rmd
worked on results, minor adjustments to table

**A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.** To link drought responss to fundamental physiological characteristics, and because measuring and modeling drought responses of every species is infeasible in diverse forests, it is important to understand how traits shape drought responses. Commonly measured traits including wood density ($WD$) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and ring porosity (Elliot et al. 2015, Friedrichs et al. 2009) have been linked to drought responses in temperate deciduous forests--as well as in other forest biomes (**REFS**). However, these traits have less direct linkage to plant hydraulic function than leaf hydraulic traits such as leaf area shrinkage upon dessication ($PLA_{dry}$; *Scoffoni et al. 2013- DOI: 10.1104/pp.113.221424*) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs ([Bartlett et al. 2016](https://www.pnas.org/content/113/46/13098.short)), which are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ ([Medeiros et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)). There is also evidence that ecosystems with high hydraulic trait diversity--but not traits like $LMA$ and $WD$-- show more modest ecosystem flux variation in response to drought ([Anderegg et al. 2018 (**but review this pub**)](https://doi.org/10.1038/s41586-018-0539-7)), but the ability of hydraulic traits to tree performance under drought remains untested. (**But see D'Orangeville et al. 2018--except that it may not be a well-constructed test for traits**)

@@

Here, we combine tree-ring records covering three droughts (1966, 1977, 1999), species functional and hydraulic trait measurments, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to test a series of hypotheses and associated specific predictions (Table 1) designed to yield functional understanding of how tree size, microenrivonment, and species' traits collectively shape drought responses. 
First, we focus on the role of tree size and its interaction with microenvironement. 
We confirm that, consistent with most forests globally, larger-diameter trees have lower drought resistance in this forest, which is in an ecoregion represented by only one study in [@bennett_larger_2015] (*H1.0*). 
We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and root water access, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*). 
Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height. We hypothesize that drought resistance will follow predicted and observed patterns in relation to wood density, specific leaf area, and ring porosity , but that leaf hydraulic traits such as leaf area shrinkage upon dehydration and turgor loss point will prove better predictors (*H2.1*). 
We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought (*H2.3*). 
Finally, we focused on variability among droughts, asking how community resistance varied across droughts (*H3.1*) and whether the factors confirming vulnerability or resistance varied across droughts (*H3.2*).

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(hypothesis_table, booktabs = TRUE, caption = "**Table 1. Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported / rejected when the response matches/contradicts the prediction in both univariate and multivariate models (when applicable). Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction in some but not all models.") %>%
  pack_rows("H1.0. Larger-diameter trees have lower drought resistance.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.1. Drought resistance decreases with tree height.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.2. Drought resistance decreases with crown exposure.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.3. Small trees (lower root volume) suffer more in drier microhabitats.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H2.1. Species traits predict drought resistance.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("H2.2. Taller trees have more drought-resistant traits.", 11, 15, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H2.3. Size-dependent drought resistance is not driven by functional traits.", 16, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H3.1. Responses varied by drought.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("H3.2. Predictor variables varied across droughts.", 18, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
  column_spec(1, width = "12cm") %>%
  column_spec(2:5, width = "1cm") %>%
  column_spec(6, width = "2cm") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%
  kableExtra::landscape()
```

@@

Community-level tree growth responses to these droughts were modest, with modal resistance values of #, #, and # for 1966, 1977, and 1999, respectively (Fig. 1b). In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($R$ $\le$ 0.7): #% in 1966, #2% un 1977, and #% in 1999. *Some* trees exhibited increased growth: ($R$ > 1.0): #% in 1966, #% un 1977, and #% in 1999. Within the context of mixed effects models, there were no significant differences in resistance across drought years (Table 4).

@@

In the non-drought years for which we have vertical profiles in climate data (**YEARS**), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig. 2). Specifically, [**add content on vertical profiles of climate variables**]

@@

Crown position was generally correlated with $H$, but with substantial variation (Fig. 2e). Crown position was a much poorer predictor of $R$ than was $H$ (Table 4), lending little overall support to *H1.2* (Table 1). When considered alone, $CP$ had a significant influence only in the 1977 drought, during which trees with dominant $CP$ had the lowest $R$. When considered in conjunction with $H$, $CP$ came out as a significant predictor only for the 1999 drought, during which suppressed trees had the lowest $R$. Crown position was included in **roughly half** of the top models, with mixed results as to how $R$ varied with $CP$ (Table 5). Most commonly in these multivariate models, as in the univariate models (Table 4), the resistance of dominant trees was less than that of co-dominant trees but higher that of suppressed trees. Thus, $CP$ was sometimes a useful predictor of $R$, but overall had a weak effect relative to that of $H$.

@@

![**Figure 2. Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity (RH), (c) air temperature, and (d) biological temperature averaged over each month of the peak growing season (May-August) from YEAR-YEAR. In these plots, heights are slightly offset for visualization purposes. Also shown are (e) 2018 tree heights by canopy position (see Table 2 for codes) and vertical profiles in (f) $PLA_{dry}$ and (g) $\pi_{tlp}$. In (f-g), values are community-wide averages across height bins (plotted at upper end of height bin), with grey indicating bins for which species-level trait measurements are available for <75% of individuals. In all plots, the dahsed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/Figure2.png){width=500px}

*Species' traits and drought resistance*
H2.1, H2.2, H2.3
Hydraulic traits alone, traits with height

**Results for third main question: are responses similar or variable across individual drought years?**
H3.1 and H3.2
Combining biophysical with hydraulic traits, which come out as candidates for best model?

>>>>

**A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.** To link drought responss to fundamental physiological characteristics, and because measuring and modeling drought responses of every species is infeasible in diverse forests, it is important to understand how traits shape drought responses. Commonly measured traits including wood density ($WD$) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and xylem architecture (Elliot et al. 2015, Friedrichs et al. 2009) have been linked to drought responses in temperate deciduous forests--as well as in other forest biomes (**REFS**). However, these traits have less direct linkage to plant hydraulic function than leaf hydraulic traits such as leaf area shrinkage upon dessication ($PLA_{dry}$; *Scoffoni et al. 2013- DOI: 10.1104/pp.113.221424*) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs ([Bartlett et al. 2016](https://www.pnas.org/content/113/46/13098.short)), which are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ ([Medeiros et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)). There is also evidence that ecosystems with high hydraulic trait diversity--but not traits like $LMA$ and $WD$-- show more modest ecosystem flux variation in response to drought ([Anderegg et al. 2018 (**but review this pub**)](https://doi.org/10.1038/s41586-018-0539-7)), but the ability of hydraulic traits to tree performance under drought remains untested. (**But see D'Orangeville et al. 2018--except that it may not be a well-constructed test for traits**)

@@

Here, we combine tree-ring records covering three droughts (1966, 1977, 1999), species functional and hydraulic trait measurments, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to test a series of hypotheses and associated specific predictions (Table 1) designed to yield functional understanding of how tree size, microenrivonment, and species' traits collectively shape drought responses. 
First, we focus on the role of tree size and its interaction with microenvironement. 
We confirm that, consistent with most forests globally, larger-diameter trees have lower drought resistance in this forest, which is in an ecoregion represented by only one study in [@bennett_larger_2015] (*H1.0*). 
We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and root water access, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*). 
Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height. We hypothesize that drought resistance will follow predicted and observed patterns in relation to wood density and specific leaf area, but that hydraulic traits including xylem architecture (*i.e.*, ring, semi-ring, or diffuse porous), leaf area shrinkage upon dehydration, and turgor loss point will prove better predictors (*H2.1*). 
We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought (*H2.3*). 
Finally, we focused on variability among droughts, asking how community resistance varied across droughts (*H3.1*) and whether the factors confirming vulnerability or resistance varied across droughts (*H3.2*).

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(hypothesis_table, booktabs = TRUE, caption = "**Table 1. Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported / rejected when the response matches/contradicts the prediction in both univariate and multivariate models (when applicable). Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction in some but not all models.") %>%
  pack_rows("H1.0. Larger-diameter trees have lower drought resistance.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.1. Drought resistance decreases with tree height.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.2. Drought resistance decreases with crown exposure.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.3. Small trees (lower root volume) suffer more in drier microhabitats.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H2.1. Species traits predict drought resistance.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("H2.2. Taller trees have more drought-resistant traits.", 11, 15, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H2.3. Size-dependent drought resistance is not driven by functional traits.", 16, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H3.1. Overall community drought resistance differed across droughts.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("H3.2. The direction of responses to predictor variables was consistent across droughts.", 18, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H3.3. The strength of responses to predictor variables varied across droughts.", 18, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
  column_spec(1, width = "12cm") %>%
  column_spec(2:5, width = "1cm") %>%
  column_spec(6, width = "2cm") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%
  kableExtra::landscape()
```

@@

Community-level tree growth responses to these droughts were modest, with modal resistance values of #, #, and # for 1966, 1977, and 1999, respectively (Fig. 1b). In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($R$ $\le$ 0.7): #% in 1966, #2% un 1977, and #% in 1999. *Some* trees exhibited increased growth: ($R$ > 1.0): #% in 1966, #% un 1977, and #% in 1999.

@@

@@

Crown position was generally correlated with $H$, but with substantial variation (Fig. 2**e**). Crown position was a much poorer predictor of $R$ than was $H$ (Table 4), lending little overall support to *H1.2* (Table 1). When considered alone, $CP$ had a significant influence only in the 1977 drought, during which trees with dominant $CP$ had the lowest $R$. When considered in conjunction with $H$, $CP$ came out as a significant predictor only for the 1999 drought, during which suppressed trees had the lowest $R$. Crown position was included in **roughly half** of the top models, with mixed results as to how $R$ varied with $CP$ (Table 5). Most commonly in these multivariate models, as in the univariate models (Table 4), the resistance of dominant trees was less than that of co-dominant trees but higher that of suppressed trees. Thus, $CP$ was sometimes a useful predictor of $R$, but overall had a weak effect relative to that of $H$.

In the non-drought years for which we have vertical profiles in climate data (**YEARS**), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig. 2). Specifically, maximum daily wind speeds were signficantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig. 2a). Relative humidity was also somewhat lower during June-August, ranging from ~50-80 above the canopy and ~60-90% in the understory (Fig. 2b). Air temperature did not vary across the vertical profile (Fig. 2c).  [**biological temperature?- probably drop**]

![**Figure 2. Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), (c) air temperature ($T_{air}$), and (d) biological temperature ($T_{biological}$)averaged over each month of the peak growing season (May-August) from YEAR-YEAR. In these plots, heights are slightly offset for visualization purposes. Also shown are (e) 2018 tree heights by canopy position (see Table 2 for codes) and vertical profiles in (f) $PLA_{dry}$ and (g) $\pi_{tlp}$. In (f-g), values are community-wide averages across height bins (plotted at upper end of height bin), with grey indicating bins for which species-level trait measurements are available for <75% of individuals. In all plots, the dahsed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/Figure2.png){width=500px}

@@

*Species' traits, height, and drought resistance*

We partially support *H2.1*: Species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were useful in explaining variation in drought responses, whereas $LMA$ and $WD$ were not (Tables 1,4,5). Specifically, $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.23; Table 4), and therefore these were excluded as candidate variables for the full multivariate models. In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained at least modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4). Of these, $PLA_{dry}$ was the strongest predictor, with consistently negative coefficients across all droughts. $\pi_{tlp}$ never came out as significant (dAIC $\ge$ 2) in the univariate models, but had a consistently negative coefficient (Table 4). Whereas ring-porous species had highest $R$ overall and in the 1966 and 1999 droughts, they had lower $R$ in 1977. Results were similar in the context of multivariate models (Table 5), except that $\pi_{tlp}$ had a positive coefficient in the 1966 models in which it was included.

We reject *H2.2*, finding no evidence that taller trees tend to have traits associated with greater drought resistance. In part because of the large sample size (n=# trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha **Ian, confirm**), there were very significant (p<0.0001) correlations of $H$ with all species' traits (**SI Table?**). However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the case of $LMA$, which did not correlate with $R$. Furthermore, although correlations were statistically signifcant, trait variation within each height class overwhelmed any vertical trends (Fig. 2**f-g**).

We support the hypothesis (*H2.3*) that the observed tendency for larger trees to have greater growth reductions during drought (lower $R$) is *not* driven by more drought-sensitive traits in larger trees, but rather by size itself (Tables 1,5). As discussed above, there was little meaningful variation in traits with height at the community level. When $H$ and hydraulic traits were considered together in multivariate models, the effect of $H$ on $R$ was consistently negative (Table 5)--reversing a non-signficant positive $H$-$R$ correlation in the univariate model for the 1999 drought (Table 4).

*Responses across droughts*

We reject *H3.1*. Within the context of mixed effects models, there were no significant differences in $R$ across drought years (Table 4). This is consistent with the observation that the distribution of $R$ values was similar across droughts (Fig. 1b).

We support Despite the broad similarity in the extent to which the three droughts affected the community as a whole, there was variation as to which variables best predicted drought responses (*H3.2b*), and occasionally in the direction of these responses (*H3.2b*).
